\documentclass[11pt]{article}

\usepackage{amsmath, amssymb}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage{parskip}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows.meta, positioning, fit, backgrounds, calc}

\geometry{margin=1in}

\title{\textbf{Grad AI Project Proposal}}
\author{Elanor Tang and Amar Shah}
\date{\today}

\begin{document}

\maketitle

\section{Background}

Our project aims to address the following question: \textbf{Are formal
specifications less susceptible to RL reward hacking compared to unit tests?}.
Specifically, we will look at the state-of-the-art Rust verifier
Verus~\cite{Verus}, which has already been used in projects at the intersection
of AI and verification \cite{AutoVerus,alphaverus,verussage,RAGVerus,vericode}. 

\textbf{Formal Verification.}
Formal verification is the process of mathematically proving that a program satisfies a precise logical specification.
Unlike testing, which checks correctness on a finite set of inputs, formal verification provides guarantees over all possible inputs and program behaviors.
% Rust is a systems programming language designed for memory safety and performance; its strong ownership type system eliminates entire classes of bugs (e.g., null-pointer dereferences, data races) at compile time, making it a natural fit for safety-critical software.

\textbf{Verus.}
Verus~\cite{Verus} is a deductive verifier for Rust that allows programmers to annotate their code with logical pre- and post-conditions, loop invariants, and other proof obligations.
Because Verus specifications are checkable and compositional, they serve as unambiguous descriptions of program intent---beyond what unit tests can capture.

\textbf{Reward Hacking.}
Reward hacking occurs when a model trained via reinforcement learning (RL) discovers behaviors that achieve high ``reward'' without actually satisfying the underlying intent of the task~\cite{RewardHacking}.
In code generation, a common form of reward hacking is overfitting to a test suite: the model learns to produce code that passes the provided tests while failing on edge cases not covered by those tests.
% Replacing hand-written or model-generated unit tests with formal specifications as the reward signal is a natural candidate for mitigating this failure mode, since a formal specification can capture the full intended semantics of a function.

\textbf{Propose-Solve-Verify.}
Our work will build upon Propose, Solve, Verify (PSV)~\cite{PSV}, which frames program synthesis as a self-play game over formal verification.
In PSV, one model \emph{proposes} a problem (i.e., a Verus specification), a second model \emph{solves} it by producing a Rust implementation, and the Verus verifier acts as an objective judge to \emph{verify} whether the solution is correct.
This setup enables scalable self-play without human-labeled correctness signals.
Our project aims to understand one of the hypotheses of this work: whether formal specifications are more robust reward signals than unit tests for RL-based code generation.

\section{Proposal}
 Our experimental set-up is as follows:
\begin{enumerate}
    \item We start with a benchmark of coding problems 
    defined by both natural language and formal Verus specifications. 
    We consider these formal Verus specifications to be the ``gold-standard" 
    which ultimately defines the coding problem.
    \item We provide a model with just the natural language specifications, 
    and ask it to generate 
    (A) unit tests to check if an implementation satisfies the specification and 
    (B) a formal Verus specification. 
    Note that this generated spec may differ from the original gold-standard spec.
    \item For both (A) and (B), we train a model via RL to produce code satisfying
    the natural language specification, 
    where the reward mechanism is either passing all the unit tests 
    or satisfying the generated Verus specification, as checked by Verus.
    \item We evaluate the two models by checking if the code produced also satisfies
    the gold-standard spec. 
    If it does not, we interpret this to mean that the model successfully ``hacked''
    the reward mechanism. 
    We hypothesize that we will see more cases 
    when the produced code passes the unit tests but does not satisfy the gold-standard spec,
    compared to using generated specs as the reward mechanism.
\end{enumerate}
\begin{figure}[h]
\centering
\tikzset{
  box/.style={rectangle, rounded corners=3pt, draw, align=center,
              minimum width=3.2cm, minimum height=1.0cm, font=\normalsize},
  goldbox/.style={box, fill=yellow!50, draw=orange!90!black, very thick},
  genbox/.style={box, fill=blue!30, draw=blue!80!black, very thick},
  rlbox/.style={box, fill=green!20, draw=green!60!black, very thick},
  evalbox/.style={box, fill=red!30, draw=red!80!black, very thick},
  arr/.style={-{Stealth[length=5pt]}, thick},
}
\begin{tikzpicture}[node distance=0.7cm and 1.8cm]

  % Top: benchmark input
  \node[goldbox, minimum width=5cm] (bench)
    {Benchmark\\{\small NL spec + gold Verus spec}};

  % NL spec fed to generator
  \node[rlbox, below=of bench] (gen)
    {LLM Generator\\{\small (given NL spec only)}};
  \draw[arr] (bench) -- node[right, font=\small]{NL spec} (gen);

  % Two branches
  \node[genbox, below left=1.2cm and 2.2cm of gen]  (tests)
    {Unit Tests};
  \node[genbox, below right=1.2cm and 2.2cm of gen] (spec)
    {Verus Spec\\{\small (generated)}};
  \draw[arr] (gen) -- (tests);
  \draw[arr] (gen) -- (spec);

  % RL training nodes
  \node[rlbox, below=1.1cm of tests] (rlA)
    {LLM Implementor\\{\small goal: pass unit tests}};
  \node[rlbox, below=1.1cm of spec]  (rlB)
    {LLM Implementor\\{\small goal: satisfy Verus spec}};
  \draw[arr] (tests) -- (rlA);
  \draw[arr] (spec)  -- (rlB);

  % Code outputs
  \node[box, below=0.9cm of rlA] (codeA) {Code A};
  \node[box, below=0.9cm of rlB] (codeB) {Code B};
  \draw[arr] (rlA) -- (codeA);
  \draw[arr] (rlB) -- (codeB);

  % Evaluation node
  \node[evalbox, minimum width=5cm,
        below=0.9cm of $(codeA)!0.5!(codeB)$] (eval)
    {Evaluate \\{\small reward hacking $\Leftrightarrow$ gold Verus spec not satisfied}};
  \draw[arr] (codeA) -- (eval);
  \draw[arr] (codeB) -- (eval);

  % Gold spec dashed arrow going around the right side
  \draw[arr, dashed, draw=orange!80!red!90!black, very thick] (bench.east)
        -- ++(6.0,0)
        |- (eval.east)
        node[pos=0.25, right, font=\normalsize\bfseries, text=orange!80!red!90!black]{gold spec};

\end{tikzpicture}
\caption{Experimental setup. Both branches receive only the natural language specification as input. The dashed arrow indicates the gold-standard Verus specification used solely for final evaluation.}
\label{fig:setup}
\end{figure}

Note: it's possible that the code produced by the models does satisfy the gold-standard specification,
    but Verus rejects it due to insufficient proof annotations. 
    If we find this to be the case, 
    we would add a step between (3) and (4) which asks a model to add only proof annotations to the produced code,
    to help Verus see that the implementation does satisfy the specification.

\textbf{Benchmarks}: We will evaluate this on the human eval Verus
benchmark~\cite{alphaverus}, which consists of around 100 coding problems with
both natural language specifications, formal Verus specifications, and a Rust
implementation. This is the best starting point: it is the largest known Verus
benchmark set with both natural language and formal specifications. Time
permitted, we could explore other benchmark sets.
%  such as the ones used in RAGVerus~\cite{RAGVerus}, Vericode~\cite{vericode}, or Verus SAGE~\cite{verussage}. However, all of these benchmark sets are missing one of the three aforementioned components (natural language specification, formal specification, Rust implementation), so we would need to either generate the missing component using a model or manually annotate it ourselves.

\textbf{Confounding Factors:} We identify a few confounding factors that could challenge the validity of our results:

\begin{enumerate}
    \item \textbf{Bad coverage by unit tests:} If the unit testing method is not well-designed, it may fail to capture important edge cases, making it easier for the model to hack the reward mechanism.
    %  To mitigate this, we will explore
    % existing literature for test case generation.
    \item \textbf{Data contamination:} The human-eval Verus benchmark is publicly available, so there is a risk that the models we use have been trained on this data, which could bias our results. 
    % For this course project, we accept this as a risk. In future work, we could explore generating our own benchmark set.
    \item \textbf{Specifications are an under approximation:} A model could generate correct code that does not satisfy the gold standard specification.
    \item \textbf{Models are worse at generating specifications:}. A model could struggle to generate correct specifications, but may be much more proficient
    at generating unit tests.
\end{enumerate}

\bibliographystyle{plain}
\bibliography{references}
\end{document}