\section{Background}
\label{background}

A \emph{formal specification} mathematically describes the expected properties of a program. 
Unlike testing, which checks a finite set of inputs, 
formal specifications can quantify over all possible inputs, 
enabling them to provide guarantees over all possible program behaviors.
\emph{Formal verification} is the process of mathematically proving that a program satisfies its specification.
Thus, it shifts the burden from ensuring correctness of an entire implementation 
(which may have complex performance optimizations)
to ensuring correctness of a higher-level mathematical specification, 
an easier task. 

% typically using a verification tool to ensure proof correctness.

One state-of-the-art verification tool is Verus~\cite{Verus}, a semi-automated Rust verifier which 
has already been used at the intersection
of AI and verification \cite{AutoVerus,alphaverus,verussage,RAGVerus,vericode}. 
In Verus, users can write three kinds of code: 
\begin{itemize}
    \item \emph{spec} code describes higher-level program properties such as function pre- and post-conditions
    or loop invariants.
    \item \emph{proof} annotations help give Verus hints for proving an implementation satisfies its spec:
    Verus is semi-automated, meaning it can automatically prove many specifications, 
    but when more complex reasoning is required,
    it may need some help in knowing intermediate proof goals 
    or useful lemma calls.
    \item \emph{exec} code is simply regular Rust executable code. 
    During compilation, Verus erases spec and proof code, 
    leaving only the exec code: 
    there is no runtime performance cost associated with adding Verus proofs.
\end{itemize}
Verus is designed to integrate directly into Rust code: 
it relies on the Rust type-checker and uses Rust macros and annotations.
However, using Verus does involve some Verus-specific syntax, 
such as additional mathematical keywords or naming function outputs to describe their expected properties.

% Verus will then check if the code satisfies these requirements. 
% Note that Verus's verification result is only as good as the specification quality: 
% if the specification is incorrect or fails to capture all possible cases, 
% implementations verified against such specifications will also lack full guarantees.
% And since Verus specifications are checkable and compositional, they can serve as unambiguous descriptions of program intent---beyond what unit tests can capture. 
% Verus~\cite{Verus} has already been used in projects at the intersection
% of AI and verification \cite{AutoVerus,alphaverus,verussage,RAGVerus,vericode}. 