\section{Challenges}
\label{challenges}

We encountered two significant challenges as we implemented our pipeline:
generating correct Verus code using an LLM, 
and checking if an implementation met the gold-standard Verus specification. 

\subsection{Generating Verus Code}

While LLMs are very familiar with Rust code, 
smaller LLMs have less experience with Verus-specific syntax.
Thus, they struggled to generate Verus code in multiple ways:
\begin{itemize}
    \item Generating syntactically incorrect 
    Verus specs from the NL description.
    \item Generating implementations for Verus specs that failed to parse,
    due to mistakes like using executable functions not supported by Verus
    or incorrect proof/exec syntax.
    \item Even if the spec and implementation used correct syntax,
    failing to generate the logically-correct proof annotations 
    for helping prove that the implementation satisfied the spec.
\end{itemize}
We illustrate the failure breakdown in Figure [insert], 
which shows a more detailed analysis of Figure~\ref{fig:results-pass-at-k}(b).
\etm{TODO}

We plan to address this challenge by providing more few-shot examples,
to help model correct Verus syntax,
and by giving more specific instructions in our prompt. 
% We plan to explore other benchmarks, such as Dafny2Verus~\cite{alphaverus} and MBPP-Verified~\cite{austin2021programsynthesislargelanguage}, in future work.
\etm{Maybe rephrase now that we've tried more of this, and mention how we give spec helpers}
We will also try other benchmarks
(Dafny2Verus~\cite{alphaverus} and MBPP-Verified~\cite{austin2021programsynthesislargelanguage}), 
to see if models found the HumanEval problems particularly hard.

\subsection{Proving Correctness Against the Gold-Standard}

Initially, we gave an LLM (Claude Opus 4.5) 
the generated implementation and asked it to add only proof annotations, 
to help Verus prove that the implementation satisfied the gold-standard spec;
if Claude was unable to do so, we interpreted that to mean the implementation was incorrect.
However, even when strongly prompted against this, 
Claude would occasionally modify the implementation to make it satisfy the spec.
Additionally, it is hard in general to verify an implementation written independently of the spec;
it was quite possible that the implementation was correct, 
but Claude was unable to find the necessary proof annotations. 

Our next approach was to use the gold-standard specs to generate \emph{property-based tests}
which sample from the input space determined by the Rust type, 
subject to the constraints outlined in the specs.
Since we already had a verified gold-standard implementation, 
we used that implementation to provide the expected output,
and tested if the generated implementations gave the same result. 

However, it was difficult to achieve good coverage merely by sampling from the input space:
in many cases, the target function accepted all inputs of a given type,
but only did something interesting if that input took a specific form. 
Since sampling uniformly generated these inputs with very low probability,
we needed to add task-specific instructions to our test generation script.
This is not scalable.

We are currently exploring alternate approaches:
we will try asking an LLM to generate an input on which the generated
and gold-standard implementations differ, if it exists.
% Since having a concrete counter-example only guarantees that the implementations differ 
% (as the lack of a counter-example is not a guarantee that they are the same),
% we would do this in combination with a test suite.
We will also try generating tests from the gold-standard \emph{implementation} rather than the spec,
in a way that achieves full coverage over all of its control branches.
We hope this will mitigate the coverage issues outlined above.